---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version-ago/nngeo)](https://cran.r-project.org/package=nngeo)
[![CRAN_Downloads_Badge](http://cranlogs.r-pkg.org/badges/last-month/nngeo)](https://cran.r-project.org/package=nngeo)

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# nngeo

The goal of `nngeo` is to provide **k-nearest neighbor join** capabilities for spatial analysis in R. 

## Installation

CRAN version - 

```{r installation1, eval=FALSE}
install.packages("nngeo")
```

GitHub version -

```{r installation2, eval=FALSE}
install.packages("devtools")
devtools::install_github("michaeldorman/nngeo")
```

## Usage

Once installed, the library can be loaded as follows.

```{r}
library(nngeo)
```


## Example

The following code section finds the 20-nearest towns for each city using function `st_nn`. 

```{r}
data(towns)
data(cities)
nn = st_nn(cities, towns, k = 20, progress = FALSE)
nn
```

By default, the result is a sparse list with the neighbor IDs.

The `st_nn` function can also be used as a **geometry predicate function** when performing spatial join with `sf::st_join`.

```{r}
st_join(cities, towns, join = st_nn, k = 3, progress = FALSE)
```

A helper function `st_connect` can be used to draw lines according to joined feature IDs. This may be useful for visualization or for network analysis. 

```{r, fig.align='center', fig.width=5, fig.height=5, warning=FALSE}
l = st_connect(cities, towns, ids = nn, progress = FALSE)
plot(st_geometry(towns), col = "darkgrey")
plot(st_geometry(cities), col = "red", add = TRUE)
plot(st_geometry(l), add = TRUE)
```












