---
title: "Introduction to package `nngeo`"
author: "Michael Dorman"
date: "`r Sys.Date()`"
output: 
  rmarkdown::pdf_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to package 'nngeo'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(nngeo)
```

# Introduction

## Package purpose

This document introduces the `nngeo` package. The `nngeo` package includes functions for spatial join of laters based on *k-nearest neighbor* relation between features. The functions work with spatial layer object defined in package `sf`, namely classes `sfc` and `sf`. 

## Installation

GitHub version:

``` r
install.packages("devtools")
devtools::install_github("michaeldorman/nngeo")
```

## Sample data

The `nngeo` package comes with three sample datasets:

* `cities`
* `towns`
* `water`

```{r, include=FALSE}
data(cities)
data(towns)
data(water)
```

The `cities` layer is a **point** layer representing the location of the three largest cities in Israel. 

```{r}
cities
```

The `towns` layer is another **point** layer, with the location of all large towns in Israel, compiled from a different data source:

```{r}
towns
```

The `water` layer is an example of a **polygonal** layer. This layer contains four polygons of water bodies in Israel.

```{r}
water
```

Figure \ref{fig:layers} shows the spatial configuration of the `cities`, `towns` and `water` layers.

```{r layers, fig.align='center', fig.width=5, fig.height=5, fig.cap='Visualization of the \\texttt{water}, \\texttt{towns} and \\texttt{cities} layers'}
plot(st_geometry(water), col = "lightblue")
plot(st_geometry(towns), col = "grey", pch = 1, add = TRUE)
plot(st_geometry(cities), col = "red", pch = 1, add = TRUE)
```

# Usage examples

## The `st_nn` function

The main function in the `nngeo` package is `st_nn`. 

The `st_nn` function accepts two layers, `x` and `y`, and returns a list with the same number of elements as `x` features. Each list element `i` is an integer vector with all indices `j` for which `x[i]` and `y[j]` are **nearest neighbors**. 

For example, the following expression finds which feature in `towns[1:5, ]` is the nearest neighbor to each feature in `cities`:

```{r}
nn = st_nn(cities, towns[1:5, ])
nn
```

This output tells us that `towns[4, ]` is the nearest among the five features of `towns[1:5, ]` to `cities[1, ]`, etc.

## The `st_connect` function

The resulting nearest neighbor matches can be visualized using the `st_connect` function. This function builds a line layer connecting features from two layers `x` and `y` based on the relations defined in a list such the one returned by `st_nn`:

```{r}
l = st_connect(cities, towns[1:5, ], ids = nn)
l
```

Plotting the line layer `l` gives a visual demonstration of the nearest neighbors match, as shown in Figure \ref{fig:st_connect}.

```{r st_connect, fig.align='center', fig.width=5, fig.height=5, fig.cap="Nearest neighbor match between \\texttt{cities} (in red) and \\texttt{towns[1:5, ]} (in grey)"}
plot(st_geometry(towns[1:5, ]), col = "darkgrey")
plot(st_geometry(l), add = TRUE)
plot(st_geometry(cities), col = "red", add = TRUE)
text(st_coordinates(cities)[, 1], st_coordinates(cities)[, 2], 1:3, col = "red", pos = 4)
text(st_coordinates(towns[1:5, ])[, 1], st_coordinates(towns[1:5, ])[, 2], 1:5, pos = 4)
```

## Dense matrix representation

The `st_nn` can also return the complete logical matrix indicating whether each feature in `x` is a neighbor of `y`. To get the dense matrix, instead of a list, use `sparse=FALSE`.

```{r}
nn = st_nn(cities, towns[1:5, ], sparse = FALSE)
nn
```

## k-Nearest neighbors where `k>0`

It is also possible to return any **k-nearest** neighbors, rather than just one. For example, setting `k=2` returns both the 1^st^ and 2^nd^ nearest neighbors:

```{r}
nn = st_nn(cities, towns[1:5, ], k = 2)
nn
```

```{r}
nn = st_nn(cities, towns[1:5, ], sparse = FALSE, k = 2)
nn
```

## Distance matrix

Using `returnDist=TRUE` the distances `list` is also returned, in addition the the neighbor matches, with both componenets now comprising a `list`:

```{r}
nn = st_nn(cities, towns[1:5, ], k = 2, returnDist = TRUE)
nn
```

## Search radius

Finally, the search for nearest neighbors can be limited to a **search radius** using `maxdist`. In the following example, the search radius is set to 50,000 meters (50 kilometers). Note that no neighbors are found within the search radius for `cities[2, ]`:

```{r}
nn = st_nn(cities, towns[1:5, ], k = 2, maxdist = 50000)
nn
```

## Spatial join

The `st_nn` function can also be used as a **geometry predicate function** when performing spatial join with `sf::st_join`. For example, the following expression spatially joins the two nearest `towns[1:5, ]` features to each `cities` features, using a search radius of 50 km:

```{r, results='hide'}
cities1 = st_join(cities, towns[1:5, ], join = st_nn, k = 2, maxdist = 50000)
```

Here is the resulting layer:

```{r}
cities1
```

## Another example

Here is another example, finding the 10-nearest neighbor `towns` features for each `cities` feature:

```{r, results='hide', warning=FALSE}
x = st_nn(cities, towns, k = 10)
l = st_connect(cities, towns, ids = x)
```

The result is visualized in Figure \ref{fig:cities_towns}. 

```{r cities_towns, fig.align='center', fig.width=5, fig.height=5, warning=FALSE, fig.cap="Nearest 10 \\texttt{towns} features from each \\texttt{cities} feature"}
plot(st_geometry(l))
plot(st_geometry(cities), col = "red", add = TRUE)
plot(st_geometry(towns), col = "darkgrey", add = TRUE)
```

# Polygons

Nearest neighbor search also works for non-point layers. The following code section finds the 20-nearest `towns` features for each water body in `water[-1, ]`. 

```{r}
nn = st_nn(water[-1, ], towns, k = 20)
```

Again, we can calculate the respective lines for the above result using `st_connect`. Since one of the inputs is line/polygon, we need to specify a sampling distance `dist`, which sets the resolution of connecting points on the shape exterior boundary. 

```{r, warning=FALSE}
l = st_connect(water[-1, ], towns, ids = nn, dist = 100)
```

The result is visualized in Figure \ref{fig:water_towns}.

```{r water_towns, fig.align='center', fig.width=5, fig.height=5, warning=FALSE, fig.cap="Nearest 20 \\texttt{towns} features from each \\texttt{water} polygon"}
plot(st_geometry(water[-1, ]), col = "lightblue", border = "grey")
plot(st_geometry(towns), col = "darkgrey", add = TRUE)
plot(st_geometry(l), col = "red", add = TRUE)
```




















